Counterfactual Training

### 2025-01-13 paper/presenations/slides.qmd
scatter still causing issues

---
title: Counterfactual Training
subtitle: Update Meeting Dec 2024
author: 
  - name: "**Patrick Altmeyer**"
    url: https://www.patalt.org/
  - name: Arie van Deursen
  - name: Cynthia C. S. Liem
institute: Delft University of Technology
date: today
format: 
  beamer: default
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

@@

- **Core problem**: `CounterfactualExplanation` objects are *huge*. They store `X` (the input), `y` (the output), etc. in their own memory.
- Can't easily fit `CounterfactualExplanation` objects on cluster memory and pass them around processes. 
- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint.

@@

1. Generate `nce` counterfactuals and distribute across mini-batches.
2. For each batch compute:
    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
    - Generator loss: $\ell_{\text{gen}}({x^\prime}_t,x_t;\theta)$
    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)$
3. Backpropagate all losses and update parameters.

@@

![Plausibility of faithful counterfactuals $x_{t}^\prime$ measured in terms of their distance from $x_{t}$.](www/moons_plaus.png){#fig-moons-plausib}

@@

@@

>>>>

---
title: Counterfactual Training
subtitle: Update Meeting Dec 2024
author: 
  - name: Patrick Altmeyer
    url: https://www.patalt.org/
  - name: Arie van Deursen
  - name: Cynthia C. S. Liem
institute: Delft University of Technology
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

@@

- **Core problem**: `CounterfactualExplanation` objects are *huge*. They store `X` (the input), `y` (the output), etc. in their own memory.
- Can't easily fit `CounterfactualExplanation` objects on cluster memory and pass them around processes. 
- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint (done!).
- Nested parallelization issue remains and I will not spend more time trying to make it work.

# Methodology

@@

1. Generate `nce` counterfactuals and distribute across mini-batches.
2. For each batch compute:
    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$.
    - Generator loss: $\lambda_{\text{gen}} \ell_{\text{gen}}({x^\prime}_t,x_t;\theta)$.
    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)$.
    - Regularization term for energies.
3. Backpropagate all losses and update parameters.

@@

![Plausibility of faithful counterfactuals $x_{t}^\prime$ measured in terms of their distance from $x_{t}$. Higher values indicate higher plausibility.](www/moons_plaus.png){#fig-moons-plausib}

@@

## MNIST (Vanilla vs *ECCo*)

::::{.columns}::::
:::{.column width='50%'}
![Faithful counterfactuals for conventionally trained MLP.](www/poc_model.png){#fig-mnist-model}
:::
:::{.column width='50%'}
![Faithful counterfactuals for same MLP with CT (*ECCo*).](www/poc_model_ct_ecco.png){#fig-mnist-model-ct-ecco}
:::
::::

## MNIST (*Generic* and *REVISE*)

::::{.columns}::::
:::{.column width='50%'}
![Faithful counterfactuals for same MLP with CT (*Generic*).](www/poc_model_ct_generic.png){#fig-mnist-model-ct-generic}
:::
:::{.column width='50%'}
![Faithful counterfactuals for same MLP with CT (*REVISE*).](www/poc_model_ct_model_revise.png){#fig-mnist-model-ct-model-revise}
:::
::::

@@

## Timeline

- Code base is already streamlined very well to allow for grid searches with easy configuration.
- Lots of open tasks and questions (see [issue](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/issues/8))
    1. Add MMD to measure plausibility.
    2. Evaluate adversarial robustness and generative capacity. Compare to AT, JEM.
    3. ... 
- Still aiming for ICML to commit to a timeline, but it seems unlikely to be in the shape I would like.
